`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

<!--
If you feel it necessary to include an appendix, it goes here.
-->


# Appendix I

## Initial data preparation script

```{r eval=FALSE, echo=TRUE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, pryr, openxlsx, tidyverse, data.table, DT, DescTools, RCurl, countrycode)
options(warn = -1, digits = 4, scipen = 999)
#---------------------------------------
#External data (country geocodes)
#---------------------------------------
geocodes <- fread("https://github.com/oughton/geocode/raw/master/example/result.csv") %>%
  select(country = V1, country_latitude = V2, country_longitude = V3) %>%
  mutate(ISO = countrycode(country, 'country.name', 'iso3c')) %>%
  filter(!is.na(ISO)) %>%
  select(ISO, country_latitude, country_longitude)

saveRDS(geocodes, "country_geocodes.rds")  
country_geocodes <- readRDS("country_geocodes.rds")

#---------------------------------------
#data preparation (GTD)
#---------------------------------------
tmp <- read.xlsx("data/data_preparation/globalterrorismdb_0617dist.xlsx", 
                 sheet = 1, colNames = TRUE) %>% 
  select(eventid, 
         year = iyear, 
         month = imonth, 
         day = iday, 
         country = country_txt, 
         region = region_txt, 
         provstate, 
         city, 
         latitude, # 2.7% NAs will be replaced with country level geocodes
         longitude,
         attack_type = attacktype1_txt, 
         weapon_type = weaptype1_txt, 
         target_type = targtype1_txt, 
         target_nalty= natlty1_txt, 
         group_name  = gname, 
         nkill,   # 5% NAs
         nwound,  # 9% NAs
         extended, 
         crit1_pol_eco_rel_soc = crit1, 
         crit2_publicize = crit2, 
         crit3_os_intl_hmn_law = crit3, 
         part_of_multiple_attacks = multiple, 
         attack_success = success, 
         suicide_attack = suicide, 
         individual_attack = individual,
         intl_logistical_attack = INT_LOG, 
         intl_ideological_attack = INT_IDEO 
         ) %>%
  replace_na(list(provstate = "unknown",       # replace nas with unknown
                  city =  "unknown",
                  target_nalty = "unknown")) %>%
  mutate(ISO = countrycode(country, 'country.name', 'iso3c'), # standardize country name codes
         month = if_else(month == 0, 1, month), # replace unknown month to 1 in 20 occurences 
         day = if_else(day == 0, 1, day),       # replace unknown day to 1 in 891 occurences
         date = paste(year, month, day, sep="-"),
         date = as.Date(date, format = "%Y-%m-%d"),
         weapon_type = if_else(
           weapon_type == "Vehicle (not to include vehicle-borne
                           explosives, i.e., car or truck bombs)", 
                          "Vehicle", weapon_type)) %>% # shorten lengthy name
  left_join(country_geocodes) %>% 
  mutate(latitude = ifelse(is.na(latitude), country_latitude, 
                           latitude), # replace missing lat lons with country lat lons
         longitude = ifelse(is.na(longitude), country_longitude, longitude)) %>%
  select(-c(country_latitude, country_longitude)) %>%
  # replace missing lat lons in remaining (~14) disputed/dissolved countries 
  # with country level lat long from prev obs
  mutate(latitude = if_else(is.na(latitude) & country == 
                              "People's Republic of the Congo", -0.2, latitude),
         longitude = if_else(is.na(longitude) & country == 
                               "People's Republic of the Congo", 15.8, longitude),
         latitude = if_else(is.na(latitude) & country == 
                              "Democratic Republic of the Congo", -4.0, latitude),
         longitude = if_else(is.na(longitude) & country == 
                               "Democratic Republic of the Congo", 21.7, longitude),
         latitude = if_else(is.na(latitude) & country == "North Yemen", 15.5, latitude),
         longitude = if_else(is.na(longitude) & country == "North Yemen", 48.5, longitude),
         latitude = if_else(is.na(latitude) & country == "South Yemen", 12.8, latitude),
         longitude = if_else(is.na(longitude) & country == "South Yemen", 45.0, longitude),
         latitude = if_else(is.na(latitude) & country == "Western Sahara", 27.4, latitude),
         longitude = if_else(is.na(longitude) & country == "Western Sahara", -9.0, longitude),
         latitude = if_else(is.na(latitude) & country == "Guadeloupe", 16.2, latitude),
         longitude = if_else(is.na(longitude) & country == "Guadeloupe", -61.5, longitude),
         latitude = if_else(is.na(latitude) & country == "New Caledonia", -20.9, latitude),
         longitude = if_else(is.na(longitude) & country == "New Caledonia", 165.6, longitude),
         latitude = if_else(is.na(latitude) & country == "Martinique", 14.6, latitude),
         longitude = if_else(is.na(longitude) & country == "Martinique", -61.0, longitude),
         latitude = if_else(is.na(latitude) & country == "Zaire", -2.5, latitude),
         longitude = if_else(is.na(longitude) & country == "Zaire", 28.8, longitude),
         latitude = if_else(is.na(latitude) & country == "Kosovo", 43.1, latitude),
         longitude = if_else(is.na(longitude) & country == "Kosovo", 20.7, longitude),
         latitude = if_else(is.na(latitude) & country == "Czechoslovakia", 50.6, latitude),
         longitude = if_else(is.na(longitude) & country == "Czechoslovakia", 14.0, longitude),
         latitude = if_else(is.na(latitude) & country == "Yugoslavia", 42.5, latitude),
         longitude = if_else(is.na(longitude) & country == "Yugoslavia", 20.5, longitude)
         )

#--------------------------------------------------------------
#External data (World Devlopment Indicatorsvfrom worldbank api)
#--------------------------------------------------------------
WDIsearch('conflict') # enter search text and extract code

ind = c("arms_export" = "MS.MIL.XPRT.KD",       # Arms exports (SIPRI trend indicator values)
        "arms_import" = "MS.MIL.MPRT.KD",       # Arms imports (SIPRI trend indicator values)
        "population" = "SP.POP.TOTL",           # Population, total
        "gdp_per_capita" = "NY.GDP.PCAP.KD",    # GDP per capita (constant 2010 US$)
        "refugee_origin" = "SM.POP.REFG.OR",    # Refugee population by country of origin
        "refugee_asylum" = "SM.POP.REFG",       # Refugee population by country of asylum
        "net_migration" = "SM.POP.NETM",        # Net migration
        "n_peace_keepers" = "VC.PKP.TOTL.UN",   # Presence of peace keepers 
        "conflict_index" = "IC.PI.CIR")         # conflict index (0-10)

countries_vec <- as.vector(unique(df$ISO)) # countries in gtd dataset

wdi_data <- WDI(indicator = ind, start = 1970, end = 2016, extra = TRUE) %>%
  select(year, ISO = iso3c, arms_export, arms_import, population, gdp_per_capita, 
         refugee_origin, refugee_asylum, net_migration, n_peace_keepers, conflict_index) %>% 
  drop_na(ISO) %>%
  filter(ISO %in% countries_vec) %>%
  # replacing NAs for visualization and modelling purpose
  replace_na(list(arms_export = 0, 
                  arms_import = 0, 
                  population = -1, 
                  gdp_per_capita = 0, 
                  refugee_origin = 0, 
                  refugee_asylum = 0, 
                  net_migration = 0, 
                  n_peace_keepers = 0, 
                  conflict_index = -1)) 


df <- df %>% left_join(wdi_data)
saveRDS(df, "gtd_clean_v2.rds")

# move all data to: gtd_eda/index/data  path for shiny and thesis writing
# "df" is the main file used throughout this research

#---------------------------------------
# iso3c file for worldmap
#---------------------------------------
countries <- df %>% group_by(country) %>% summarise(total = round(n())) 
countries$iso3 <- countrycode(countries$country, 
                              origin = "country.name", destination = "iso3c")
saveRDS(countries, "countries.rds")

```

## List of variables and short description

```{r echo=FALSE}
tmp <- as.data.frame(names(df))
tmp$description <- c("a 12-digit Event ID", 
           "year in which the incident occurred", 
           "month", "day", "country", "world region", 
           "an administrative division or unit of a country", 
           "city", "latitude", "longitude",  
           "method of attack (reflects the broad class of tactics used)", 
           "type of weapon used in the incident", 
           "type of target/victim", 
           "nationality of the target that was attacked", 
           "name of the group that carried out the attack", 
           "number of total confirmed fatalities for the incident", 
           "number of confirmed non-fatal injuries", 
           "whether or not an incident extended more than 24 hours", 
           "political, economic, religious, or social goal", 
           "intention to coerce, or publicize to larger audience", 
           "action from the incident is outside intl humanitarian law", 
           "whether an incident being part of multiple attacks", 
           "suicide attack",
           "whether an incident was successful", 
           "whether an attack carried out by unaffiliated Individual(s)",  
           "cross border incident", 
           "attack on target of a different nationality", 
           "ISO code for country", 
           "Approx. date of incident", 
           "Arms exports (SIPRI trend indicator values)", 
           "Arms imports (SIPRI trend indicator values)", 
           "Population, total", 
           "GDP per capita (constant 2010 US$)", 
           "Refugee population by country or territory of origin", 
           "Refugee population by country or territory of asylum", 
           "Net migration", 
           "Presence of peace keepers", 
           "Extent of conflict of interest regulation index (0-10)")
names(tmp) <- c("Name of the Variable", "description")
knitr::kable(tmp, caption = "Short description of important variables", booktabs = TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 12, full_width = F, position = "left")

```

# Appendix II

This appendix includes all of the R chunks of code that were hidden throughout the document (using the `include = FALSE` chunk tag) to help with readibility and/or setup.

**In the main Rmd file**

```{r ref.label='include_packages', results='hide', echo = TRUE}
```

**In Chapter \@ref(ref-labels):**

```{r ref.label='include_packages_2', results='hide', echo = TRUE}
```